RationalGRL: A Framework for Argumentation and Goal Modeling

### 2017-08-21 abstract.tex
Small edits and few comments abstract and intro

\begin{abstract}
Goal modeling languages capture the relations between an information system and its environment using high-level goals and their relationships with lower level goals and tasks. The process of constructing a goal model usually involves discussions between a requirements engineer and a group of stakeholders. While it is possible to capture part of this discussion process in a goal model, for instance by specifying alternative solutions for a goal, not all of the arguments can be found in the resulting model. For instance, reasons for accepting or rejecting an element or a relation between two elements are not captured. In this paper, we investigate to what extent argumentation techniques from artificial intelligence can be applied to a goal modelingn approach. We apply the argument scheme for practical reasoning (PRAS), which is used in AI to reason about goals to the Goal-oriented Requirements Language (GRL). We develop a formal metamodel for the new language, link it to the GRL metamodel, and we implement our extension into jUCMNav, the Eclipse-based open source tool for GRL.
\end{abstract}
%SG: what are the findings? or what case study we are using? 
%SG_New: I still have these comments!

>>>>

\begin{abstract}
Goal modeling languages capture the relations between an information system and its environment using high-level goals and their relationships with lower level goals and tasks. The process of constructing a goal model usually involves discussions between a requirements engineer and a group of stakeholders. While it is possible to capture part of this discussion process in a goal model, for instance by specifying alternative solutions for a goal, not all of the arguments can be found in the resulting model. For instance, the discussion on whether to accept or reject a certain goal and the ultimate rationale for acceptance or rejection cannot be captured in current goal modeling languages. Based on a case study in which stakeholders discuss requirements for a Traffic Simulator, we apply argumentation techniques from artificial intelligence to a goal modeling approach. Thus, we combine a traditional goal modelling approach, the Goal-oriented Requirements Language (GRL), with a formal Practical Reasoning Argument Scheme (PRAS) for reasoning about goals into a new framework (RationalGRL). RationalGRL provides a methodology, formal semantics and tool support to capture the discussions and outcomes of the argumentation process that leads to a goal model. 
\end{abstract}

### 2017-08-21 introduction.tex
Small edits and few comments abstract and intro

\subsection{Requirements Engineering and Goal Modeling}

Requirements Engineering (RE) is an approach to assess the role of a future information system within a human or automated environment. An important goal in RE is to produce a consistent and comprehensive set of requirements covering different aspects of the system, such as general functional requirements, operational environment constraints, and so-called non-functional requirements such as security and performance.

One of the first activities in RE are the ``early-phase'' requirements engineering activities, which include those that consider how the intended system should meet organizational goals, why it is needed, what alternatives may exist, what implications of the alternatives are for different stakeholders, and how the interests and concerns of stakeholders might be addressed~\cite{yu1997towards}. These activities fall under the umbrella of goal modeling. The are a large number of established RE methods using goal models in the early stage of requirements analysis (e.g.,~\cite{liu2004designing,donzelli2004goal,dardenne1993goal,chung2012non,castro2002towards}, overviews can be found in~\cite{van2001goal,kavakliL05}). Several goal modeling languages have been developed in the last two decades as well. The most popular ones include $i*$~\cite{Yu:1997:TMR:827255.827807}, Keep All Objects Satisfied (KAOS)~\cite{van2008requirements}, the NFR framework~\cite{chung2012non}, \textsc{Tropos}~\cite{giorgini2005goal}, the Business Intelligence Model (BIM)~\cite{horkoff2014strategic}, and the Goal-oriented Requirements Language (GRL)~\cite{Amyot:2010:EGM:1841349.1841356}.

@@

A goal model is often the result of a discussion process between a group of stakeholders. For small-sized systems, goal models are usually constructed in a short amount of time, involving stakeholders with a similar background. Therefore, it is often not necessary to record all of the details of the discussion process that led to the final goal model. However, most real-world information systems -- e.g., air-traffic management, industrial production processes, or government and healthcare services -- are complex and are not constructed in a short amount of time, but rather over the course of several workshops. In such situations, failing to record the discussions underlying a goal model in a structured manner may harm the success of the RE phase of system development for various reasons stated below.

First, it is well-known that stakeholders' preferences are rarely absolute, relevant, stable, or consistent~\cite{march1978bounded}. Therefore, it is possible that a stakeholder changes his or her opinion about a modeling decision in between two goal modeling sessions, which may require revisions of the goal model. If previous preferences and opinions are not stored explicitly, it is not straightforward to remind stakeholders of their previous opinions which results in having unnecessary discussions and revisions. As the number of participants increases, revising the goal model based on changing preferences can take up a significant amount of time.

@@

Third, alternative and different ideas and opposing views that could potentially have led to different goal diagrams could be lost. For instance, a group of stakeholders specifying a goal model for a user interface may decide to reduce softgoals ``easy to use'' and ``fast'' to one softgoal ``easy to use''. Thus, the resulting goal model merely contains the softgoal ``easy to use'', but the discussion as well as the decision to reject ``fast'' are lost. This leads to a poor understanding of the problem and solution domain. In fact, empirical data suggest that this is an important reason of RE project failure~\cite{curtis1988field}.

Fourth and lastly, in goal models in general, ``rationale'' behind any decision is static and does not immediately impact the goal models when they change. That is, current goal modeling languages have limited support for reasoning about changing beliefs and opinions, and their effect on the goal model. A stakeholder may change his or her opinion, but it is not always directly clear what its effect is on the goal model. Similarly, a part of the goal model may change, but it is not possible to formally reason about the consistency of this new goal model with the underlying beliefs and arguments in the goal modeling languages. This becomes more problematic if the participants constructing the goal model change, since modeling decisions made by one group of stakeholders may conflict with the underlying beliefs put forward by another group of stakeholders.

@@

The aim of this research is to resolve the above issues by developing a framework with tool-support, which combines goal modeling approaches with argumentation techniques from Artificial Intelligence (AI) research~\cite{atkinson2007}. We have identified five important requirements for our framework: 
\begin{enumerate}
\item The argumentation techniques should be close to the actual discussions of stakeholders or designers in the early requirements engineering phase.
\item 
The framework must have formal traceability links between elements of the goal model and underlying arguments.
\item 
Using these traceability links, it must be possible to compute the effect of changes in the goal model on the underlying arguments, and vice versa.
\item 
The framework must have a tool support.
\item 
There should be a methodology for the framework to guide the practitioners in its application in real cases.
\end{enumerate}

@@

\begin{quote}
\textbf{RQ.} What are the constructs, mechanisms and rules for developing a framework that formally captures the discussions between stakeholders such that it can generate goal models?
\end{quote}

To answer the research question above and solve some of the concerns mentioned in Subsection~\ref{sect:problem_statement}, we propose a framework called RationalGRL. The RationalGRL framework combines the Goal-oriented Requirements Language (GRL) with a technique from argumentation and discourse modeling called \emph{argument schemes}~\cite{walton-etal2004}.

In past, we introduced our preliminary steps towards RationalGRL framework~\cite{bnaic2014,vanzee-etal:renext2015,vanZee-etal:er2016} with its tool-support~\cite{vanZee-etal:comma2016}.  In our previous work, argument diagrams are integrated with GRL models to allow better representation of stakeholder's arguments and discussions. In our current paper, we improve RationalGRL framework by adding argument schemes and critical questions to argumentation framework and provide traceability links between the argumentation framework and GRL.  For this, we modify and extend our metamodel to specify how our framework extends GRL, and we use this as the basis for a prototype web-based implementation~\footnote{The tool is available at <GITHUB LINK>}. %SG: Placeholder for the link. 
We develop an initial list of argument schemes and critical questions by analyzing a set of transcripts containing discussion about an information system. Our framework is fully extensible, meaning that argument schemes and critical questions can be added by users for specific problem domains. We develop a methodology for using RationalGRL, which consists of developing goal models and posing arguments in an integrated way. This methodology is intended to be used by practitioners in the field. We formalize RationalGRL using propositional logic, and develop algorithms for adding goal elements and arguments. We illustrates the steps of our framework  and how the framework can be applied in real cases with a traffic simulator example.

@@

This article is organized as follows. Section~\ref{sect:background} contains background and introduces our running example, the Goal-oriented Requirements Language (GRL)~\cite{Amyot:2010:EGM:1841349.1841356}, and argument schemes. Section~\ref{sect:overview} provides a brief and high-level overview of our framework, together with a metamodel and the methodology. Section~\ref{sect:gmas} contains an in depth explanation of how we obtained an initial set of argument schemes by annotating transcripts from discussions about an information system, and in Section~\ref{sect:examples} we provide several examples of these arguments schemes. In Section~\ref{sect:formalframework} we formalize RationalGRL and our arguments schemes using propositional logic and we use well-known argumentation semantics to compute which arguments are accepted and which are rejected. We also develop various algorithms for the argument schemes in this section. In Section~\ref{sect:tool} we provide a brief overview of the prototype tool we developed for RationalGRL. Finally, Section~\ref{sect:discussion} contains a discussion, covering related work, future work, and a conclusion.

>>>>

\subsection{Requirements Engineering and Goal Modeling}
\todo{Floris}{all}{Starting section 1 directly with subsection 1.1 is generally frowned upon. Maybe we can get rid of the subsections in the intro?} 
Requirements Engineering (RE) is an approach to assess the role of a future information system within a human or automated environment. An important goal in RE is to produce a consistent and comprehensive set of requirements covering different aspects of the system, such as general functional requirements, operational environment constraints, and so-called non-functional requirements such as security and performance.

Among the initial activities in RE are the ``early-phase'' requirements engineering activities, which include those that consider how the intended system should meet organizational goals, why it is needed, what alternatives may exist, what the implications of the alternatives are for different stakeholders, and how the interests and concerns of stakeholders might be addressed~\cite{yu1997towards}. These activities fall under the umbrella of goal modeling. There are a large number of established RE methods using goal models in the early stage of requirements analysis (e.g.,~\cite{liu2004designing,donzelli2004goal,dardenne1993goal,chung2012non,castro2002towards}, overviews can be found in~\cite{van2001goal,kavakliL05}). Several goal modeling languages have been developed in the last two decades as well. The most popular ones include $i*$~\cite{Yu:1997:TMR:827255.827807}, Keep All Objects Satisfied (KAOS)~\cite{van2008requirements}, the NFR framework~\cite{chung2012non}, \textsc{Tropos}~\cite{giorgini2005goal}, the Business Intelligence Model (BIM)~\cite{horkoff2014strategic}, and the Goal-oriented Requirements Language (GRL)~\cite{Amyot:2010:EGM:1841349.1841356}.

@@

A goal model is often the result of a discussion process between a group of stakeholders. For small-sized systems, goal models are usually constructed in a short amount of time, involving stakeholders with a similar background. Therefore, it is often not necessary to record all of the details of the discussion process that led to the final goal model. However, goal models for many complex, real-world information systems -- e.g., air-traffic management systems, systems that support industrial production processes, or government and healthcare services -- are not constructed in a short amount of time, but rather over the course of several workshops with stakeholders and requirements engineers. In such situations, failing to record the discussions underlying a goal model in a structured manner may harm the success of the RE phase of system development for various reasons.

First, it is well-known that stakeholders' preferences are rarely absolute, relevant, stable, or consistent~\cite{march1978bounded}. Therefore, it is possible that a stakeholder changes his or her opinion about a modeling decision in between two goal modeling sessions, which may require revisions of the goal model. If previous preferences and opinions are not stored explicitly, it is not straightforward to remind stakeholders of their previous opinions which can result in unnecessary discussions and revisions. As the number of participants increases, revising the goal model based on changing preferences can take up a significant amount of time.

@@

Third, alternative and different ideas and opposing views that could potentially have led to different goal models could be lost. For instance, a group of stakeholders specifying a goal model for a user interface may decide to reduce the two goals ``easy to use'' and ``fast'' to one goal ``easy to use''. Thus, the resulting goal model merely contains the goal ``easy to use'', but the discussion as well as the decision to reject the goal ``fast'' are lost. This leads to a poor understanding of the problem and solution domain. In fact, empirical data suggest that this is an important reason of RE project failure~\cite{curtis1988field}.

Fourth and finally, in goal models in general the rationale behind any modeling decision is static and changing these rationales does not immediately impact the goal models. That is, current goal modeling languages have limited support for reasoning about changing beliefs and opinions, and their effect on the goal model. A stakeholder may change his or her opinion, but it is not always directly clear what its effect is on the goal model. Similarly, with existing goal modelling languages one can change a part of the goal model, but it is not possible to reason about whether or not this new goal model is consistent with the underlying beliefs and arguments. This becomes more problematic if the participants constructing the goal model change, since modeling decisions made by one group of stakeholders may conflict with the underlying beliefs put forward by another group of stakeholders.

\todo{Floris}{Floris}{I think the above 4 reasons overlap in some cases. If it's OK, I'll rewrite this.}

@@

The aim of this research is to resolve the above issues by developing a framework with tool-support, which combines goal modeling approaches with argumentation techniques from Artificial Intelligence (AI) research~\cite{atkinson2007}. We have identified five important requirements for our framework: 
\begin{enumerate}
\item The argumentation techniques should be close to the actual discussions of stakeholders or designers in the early requirements engineering phase.
\item 
The framework must have formal traceability links between elements of the goal model and underlying arguments.
\item 
Using these traceability links, it must be possible to compute the effect of changes in the goal model on the underlying arguments, and vice versa.
\item 
The framework must have software tool support.
\item 
There should be a methodology for the framework to guide the practitioners in its application in real cases.
\end{enumerate}

@@

\begin{quote}
\textbf{RQ.} What are the constructs, mechanisms and rules for developing a framework that formally captures the discussions between stakeholders such that it can generate goal models?
\end{quote}
\todo{Floris}{Marc,Sepideh}{While I like an explicit research question, you introduce new terminology here ``construct, mechanisms and rules''. Do you explicitly refer to this later in the paper (e.g. we found mechanisms A, B, rules R1, R2, etc.)?}
To answer the research question above and solve some of the concerns mentioned in Subsection~\ref{sect:problem_statement}, we propose a framework called RationalGRL. The RationalGRL framework combines the Goal-oriented Requirements Language (GRL) with a technique from argumentation theory and discourse modeling called \emph{argument schemes}~\cite{walton-etal2004}. \todo{Floris}{Floris}{mention Singh et al. argument schemes in RE here.}

In previous work, we have taken preliminary steps towards the RationalGRL framework~\cite{bnaic2014,vanzee-etal:renext2015,vanZee-etal:er2016} with its tool-support~\cite{vanZee-etal:comma2016}.  In our previous work, argument diagrams are integrated with GRL models to allow better representation of stakeholders' arguments and discussions. In our current paper, we improve the RationalGRL framework by adding argument schemes and critical questions to argumentation framework and provide traceability links between the argumentation framework and GRL.  For this, we modify and extend our metamodel to specify how our framework extends GRL, and we use this as the basis for a prototype web-based implementation\footnote{The tool is available at <GITHUB LINK>}. %SG: Placeholder for the link. 
Inspired by work on practical reasoning in Artificial Intelligence~\cite{atkinson2007}, we develop an initial list of argument schemes and critical questions for reasoning about goals in RE. We do this by analyzing a set of transcripts containing discussions about a traffic simulator information system. Our framework is fully extensible, meaning that argument schemes and critical questions can be added by users for specific problem domains. We develop a methodology for using RationalGRL, which consists of developing goal models and posing arguments in an integrated way. This methodology is intended to be used by practitioners in the field. We provide formal semantics for RationalGRL by translating the elements of the GRL language into simple propositional logical formulas and subsequently capturing arguments about these elements of a GRL model in a formal argumentation framework \cite{dung1995}. We then develop a set of algorithms for applying argument schemes and critical questions about goal models. These algorithms further formally ground the RationalGRL methodology and form the basis of the RationalGRL tool. We illustrate the steps of our framework and how RationalGRL can be applied in a real case with the traffic simulator case study.

@@

This article is organized as follows. Section~\ref{sect:background} introduces our running example, the Goal-oriented Requirements Language (GRL)~\cite{Amyot:2010:EGM:1841349.1841356}, the Practical Reasoning Argument Scheme (PRAS)~\cite{atkinson2007} and some general background about . Section~\ref{sect:overview} provides a brief and high-level overview of our framework, together with a metamodel and the methodology. Section~\ref{sect:gmas} contains an in depth explanation of how we obtained an initial set of argument schemes and critical questions by annotating transcripts from discussions about an information system, and in Section~\ref{sect:examples} we provide several examples of these schemes and questions. In Section~\ref{sect:formalframework} we provide formal semantics for GRL and show how argumentation semantics \cite{dung1995} can be used to compute which arguments are accepted and which are rejected. We also develop various algorithms for the argument schemes in this section. In Section~\ref{sect:tool} we provide a brief overview of the prototype tool we developed for RationalGRL. Finally, Section~\ref{sect:discussion} contains a discussion, covering related work, future work, and a conclusion.
